A distance metric between histograms through the the Kolmogorov-Smirnov test statistic: specification, measures reference and example uses

### 2015-10-31 paper.tex
finalizada versao inicial

\begin{abstract}
This document presents reference values for a distance metric
derived from the Kolmogorov-Smirnov test statistic $D_{n,n'}$.
Each measure of $D_{n,n'}$ is a distance between two histograms,
which is normalized by the number of observations in each sample
to yield p-values $c'$, i.e. values for which 
higher levels of significance $\alpha>c'$ implies the rejection of the null hypothesis.
Benchmarks for the implementation are delivered by comparing samples from known distributions.
Pattern examples in real data enables further
insight in the robustness and power of $c'$.
\end{abstract}

\pacs{05.10-a,}% PACS, the Physics and Astronomy
\keywords{Kolmogorov-Smirnov test, statistic, benchmark, distance measure, histogram}
\maketitle
\section{Introduction}\label{sec:intro}

@@

Hight values of $c$ favor rejecting the null hypothesis.
For example, if the significance level is $\alpha=0.01$,
then $c$ greater than $1.7$
implies the rejection of the null hypothesis and
suggests that $F$ and $F'$
are outcomes of different distributions.
Of core importance in this study is to regard the $c$ statistic
as a measure of distance between both distributions~\cite{kolm}.
The main contribution of the following sections is the
explicit display of reference values of $c$
from which one might derive knowledge from
$c$ measures or even from a single value of $c$.

@@

\section{Text}
This section exemplifies the use of $c$
in the detection of similarity between texts.
Each text $X$ was divided in two halves $X1$ and $X2$.
The set of known English words were considered as were 
the set of stopwords (words with reduced meaning such
as prepositions and articles).
Only the number of letters in each words was measured.
Three approaches were chosen: 1) the text was partitioned into 1000 pieces of equal number of characters, the mean of the word size of each piece is an element of the sample; 2) the text was partitioned into 1000 pieces of equal number of characters, the standard deviation of the word size is an element of the sample; 3) each word size is an element of the sample.
This last case yields a discrete probability distribution, which was approximated as a continuous variable and gave the greatest sensibility to text differences.
The overall result is the same: smaller differences between parts
of the same text.
Notice that the $c$ is often high within a same book.

@@

\FloatBarrier
\section{Audio}
This section presents $c$ values
drawn from audio for testing the sound system of the computer.
The PCM samples of the files were normalized to fit the interval
$[-1,1]$ to yield the samples labeled. The wavelet decomposition was performed with the Daubechies 8 Wavelet function.
The resulting values of the $c$ statistic reflect most of all the
different types of signals analysed:
PCM samples and wavelet decomposition coefficients in different leafs.
Among each type of signal, the type of sound is also reflected in the measures of $c$, with the noise having the highest values.

@@

\FloatBarrier
\section{Music}
This section presents measures of the $c$ statistic drawn
from the pitches of the notes of classical compositions.
The results reflect music history.
For example, measures of $c$ involving Palestrina
increases with the exception of Beethoven
who, indeed, used modalism.
The values of $c$ related to Bach also increases along time,
and the outcome of the comparison against Palestrina 
is only exceeded when Sch\"onberg is reached,
which reflects the non-tonal discourse of both
Palestrina and Sch\"onberg.

@@

\FloatBarrier
\section{OS status}
This last example expose the statistic $c$ for samples drawn from
the operational system of my laptop.
The patterns are less neat than on last examples,
but many conclusions can still be reached.
The memory used by the most consuming processes compose
the samples which present the highest values of $c$.
Lowest values of $c$ are related to RAM usage.
Again, the type of samples are mandatory:
they might all be identified by the values of $c$
found in comparison to other samples,
with the exception of the RAM memory.
\input{tables/osGeneral_}
\input{tables/osDistances_}

\FloatBarrier
\section{Conclusions and automation of the benchmarks}\label{sec:conc}
The $c$ statistic is robust both to determine if
the distributions underlying the samples are the same
and to quantify the difference between such probability distributions.
The benchmarks for $c$, given in Section~\ref{sec:simulations},
are very useful as references to make sense of data 
e.g. as the example analyses of Section~\ref{sec:empirical}.
Notice that the calculations require no
training or clusterization usually involved in
classification routines.

The rendering of this article and all tables
are automated through Python scripts
in order to ease the scrutinization of different settings~\cite{gmanePack}.
After some exploration of  the results,
the setting of
$N_c=100$ comparisons per measure,
$n=n'=1000$ elements in each sample,
$N_b=30$ equally spaced bins was settled
as the template of this document.
The main reason for this choice is that
the results are consistent and stable,
but are still of very modest scales.
The use of more bins 
should enhance the results with respect to generality.
On the other hand, samples are often not so big,
which favors reporting the tables with a small sample size.

>>>>

\begin{abstract}
This document presents reference values for a distance metric
derived from the Kolmogorov-Smirnov test statistic $D_{F,F'}$.
Each measure of $D_{F,F'}$ is a distance between two histograms,
which is normalized by the number of observations in each sample
to yield the $c$ statistic, which can be mapped to p-values, i.e. values for which 
higher levels of significance $\alpha>c(\alpha)$ implies the rejection of the null hypothesis.
Benchmarks for the implementation are delivered by comparing samples from known distributions.
Pattern examples in real data enables further
insight in the robustness and power of $c$.
\end{abstract}

\pacs{05.10-a,}% PACS, the Physics and Astronomy
\keywords{Kolmogorov-Smirnov test, statistic, benchmark, distance measure, histogram}
\maketitle

\tableofcontents

\section{Introduction}\label{sec:intro}

@@

Hight values of $c$ favor rejecting the null hypothesis.
For example, if the significance level is $\alpha=0.01$,
then $c$ greater than $1.7$
implies the rejection of the null hypothesis and
suggests that $F$ and $F'$
are outcomes of different distributions.
Of core importance in this study is to regard the $c$ statistic
as a measure of distance between both distributions~\cite{kolm}.
The main contribution of the following sections is the
explicit display of reference values of $c$
from which one might derive knowledge from
measures or even from a single value of $c$.

%\subsection{Workarounds}
%The Kolmogorov-Smirnov test is robust,
%but data often fall into the cases where
%it is not guaranteed to be valid.
%In order to make the example analysis,
%the empirical density functions are built from discrete data.
%
%as we fall into two of these.
%
% z-score
% discrete
% formulação com max(f-x,x-f2)

%If the number of elements ($n$ and $n'$)
%in each sample are bellow 50.

@@

\subsection{Text}
This section exemplifies the use of $c$
in the detection of similarity between texts.
Each text $X$ was divided in two halves $X1$ and $X2$.
The set of known English words were considered as were 
the set of stopwords (words with reduced meaning such
as prepositions and articles).
Only the number of letters in each words was measured.
Three approaches were chosen: 1) the text was partitioned into 1000 pieces of equal number of characters, the mean of the word size of each piece is an element of the sample; 2) the text was partitioned into 1000 pieces of equal number of characters, the standard deviation of the word size is an element of the sample; 3) each word size is an element of the sample.
This last case yields a discrete probability distribution, which was approximated as a continuous variable and gave the greatest sensibility to text differences.
The overall result is the same: smaller differences between parts
of the same text.
Notice that the $c$ is often high within a same book.

@@

\FloatBarrier
\subsection{Audio}
This section presents $c$ values
drawn from audio for testing the sound system of the computer.
The PCM samples of the files were normalized to fit the interval
$[-1,1]$ to yield the samples labeled. The wavelet decomposition was performed with the Daubechies 8 Wavelet function.
The resulting values of the $c$ statistic reflect most of all the
different types of signals analysed:
PCM samples and wavelet decomposition coefficients in different leafs.
Among each type of signal, the type of sound is also reflected in the measures of $c$, with the noise having the highest values.

@@

\FloatBarrier
\subsection{Music}
This section presents measures of the $c$ statistic drawn
from the pitches of the notes of classical compositions.
The results reflect music history.
For example, measures of $c$ involving Palestrina
increases with the exception of Beethoven
who, indeed, used modalism.
The values of $c$ related to Bach also increases along time,
and the outcome of the comparison against Palestrina 
is only exceeded when Sch\"onberg is reached,
which reflects the non-tonal discourse of both
Palestrina and Sch\"onberg.

@@

\FloatBarrier
\subsection{OS status}
This last example expose the statistic $c$ for samples drawn from
the operational system of my laptop.
The patterns are less neat than on last examples,
but many conclusions can still be reached.
The memory used by the most consuming processes compose
the samples which present the highest values of $c$.
Lowest values of $c$ are related to RAM usage.
Again, the type of samples are mandatory:
they might all be identified by the values of $c$
found in comparison to other samples,
with the exception of the RAM memory.
\input{tables/osGeneral_}
\input{tables/osDistances_}

\FloatBarrier
\section{Conclusions and further benchmarks}\label{sec:conc}
The $c$ statistic is robust both to determine if
the distributions underlying the samples are the same
and to quantify the difference between such probability distributions.
The benchmarks for $c$, given in Section~\ref{sec:simulations},
are very useful as references to make sense of data 
e.g. as the example analyses of Section~\ref{sec:empirical}.
Notice that the calculations require no
training or clusterization usually involved in
classification routines.

The rendering of this article and all tables
are automated through Python scripts
in order to ease the scrutinization of different settings~\cite{gmanePack}.
After some exploration of  the results,
this setting was settled as template for this document:
simulations used $N_c=100$ comparisons per measure with
$n=n'=1000$ elements in each sample;
all empirical density functions
derived from histograms with 
$N_b=30$ equally spaced bins.
The main reason for this choice is that
the results are consistent and stable,
but are still of very modest scales.
The use of more bins 
should enhance the results with respect to generality.
On the other hand, samples are often not so big,
which favors reporting the tables with a small sample size.
